Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-13 mr-review.Rmd
added count and survival outcome methods

The second type of horizontal pleiotropy, from above, is sometimes called "correlated horizontal pleiotropy". Correlated horizontal pleiotropy occurs when 
CAUSE and MRAID are two methods that aim to model and to account for correlated horizontal pleiotropy. 
Their approaches to modeling correlated horizontal pleiotropy differ in important ways that, ultimately, 
affect performance. 
CAUSE allows for a proportion of variants to exhibit correlated pleiotropy, which @morrison_mendelian_2020 
models as an effect on a shared, unobserved factor. The remaining variants are independent of this 
unobserved factor. Every variant can have a nonzero pleiotropic effect on the outcome, and these 
pleiotropic effects are uncorrelated with the variant effects on the exposure. The model, then, 
is written as a mixture of biavariate normal distributions, and inference proceeds by borrowing ideas on 
adaptive shrinkage from @stephens_false_2016. Finally, @morrison_mendelian_2020 compares two models 
to determine whether the GWAS summary statistics are consistent with a causal effect of the
exposure on the outcome. Specifically, they estimate the difference in the expected log pointwise 
posterior density for the model with causal effect fixed at zero and the model that permits a nonzero 
causal effect [@vehtari_practical_2017].

@@

Binary, survival, count - see discussion section of MRAID paper.
@allman_novel_2021 @burgess_improving_2012

While we've focused on continuous outcomes and continuous exposures, some
data are more naturally treated as binary, count, or time-to-event variables. MR methods 
for non-continuous outcomes and exposures is an active area of research. @burgess_improving_2012 and @allman_novel_2021 
developed MR strategies for binary exposures and binary outcomes. @burgess_improving_2012 extended 
methods from @wald_fitting_1940 by fitting two logistic regressions, one for the outcome and one for 
the exposure. They then calculated the causal effect estimator as the ratio of the two logistic 
regression coefficient estimators. @allman_novel_2021 devised a new MR method for binary outcome and 
binary exposure by drawing on @gauvreau_analysis_1997 and treating the exposure and outcome as correlated binary variables. Writing the likelihood for this model, they implemented an iterative optimization algorithm to simultaneously infer the causal effect parameter and other parameters.

@@

# Software

>>>>

The second type of horizontal pleiotropy, from above, is sometimes called "correlated horizontal pleiotropy". Correlated horizontal pleiotropy occurs when

CAUSE and MRAID are two methods that aim to model and to account for correlated horizontal pleiotropy. 
Their approaches to modeling correlated horizontal pleiotropy differ in important ways that, ultimately, 
affect performance. 
CAUSE allows for a proportion of variants to exhibit correlated pleiotropy, which @morrison_mendelian_2020 
models as an effect on a shared, unobserved factor. The remaining variants are independent of this 
unobserved factor. Every variant can have a nonzero pleiotropic effect on the outcome, and these 
pleiotropic effects are uncorrelated with the variant effects on the exposure. The model, then, 
is written as a mixture of biavariate normal distributions, and inference proceeds by borrowing ideas on 
adaptive shrinkage from @stephens_false_2016. Finally, @morrison_mendelian_2020 compares two models 
to determine whether the GWAS summary statistics are consistent with a causal effect of the
exposure on the outcome. Specifically, they estimate the difference in the expected log pointwise 
posterior density for the model with causal effect fixed at zero and the model that permits a nonzero 
causal effect [@vehtari_practical_2017].

@@

While we've focused above on continuous outcomes and continuous exposures, some
data are more naturally treated as binary, count, or time-to-event variables. MR methods 
for non-continuous outcomes and exposures is an active area of research. @burgess_improving_2012 and @allman_novel_2021 
developed MR strategies for binary exposures and binary outcomes. @burgess_improving_2012 extended 
methods from @wald_fitting_1940 by fitting two logistic regressions, one for the outcome and one for 
the exposure. They then calculated the causal effect estimator as the ratio of the two logistic 
regression coefficient estimators. @allman_novel_2021 devised a new MR method for binary outcome and 
binary exposure by drawing on @gauvreau_analysis_1997 and treating the exposure and outcome as correlated binary 
random variables. They implemented an iterative optimization algorithm to 
simultaneously infer the causal effect parameter and other parameters.

Recently, multiple teams of investigators have analyzed time-to-event traits in MR studies [@burgess_commentary_2015;@arvanitis_linear_2021;@tikkanen_body_2019;@he_causal_2018;@guo_body_2017]. 
@guo_body_2017, for example, studied post-diagnosis survival time in women with breast cancer. They 
treated survival as a time-to-event outcome, which enabled them to model it with a proportional 
hazards model [@cox_regression_1972] in two-stage MR with individual-level data. Similarly, 
@he_causal_2018, in a study of cardiovascular disease, applied proportional hazards regression when 
examining time to stroke as an outcome variable. We are unaware of MR studies that treat 
the exposure variable as a time-to-event, but it's conceivable that future research will consider 
time-to-event traits as exposures in MR analyses.

Researchers have also drawn on other generalized linear models for analysis in MR 
studies. However, this area is relatively unexplored. @hazewinkel_mendelian_2022 treated hospitalizations as a count variable and 
modeled it with quasi-Poisson methods. @allman_mendelian_2022 characterized statistical properties, including bias and 
power, with simulations of count random variables and binary random variables. With growing interest
in modeling counts in biomedical data, where molecular phenotyping technologies now acquire RNA molecule counts and protein abundances, 
there may be opportunities for methods innovations in a generalized linear models framework for MR studies.

See also @tchetgen_tchetgen_genius_2021

@@
